---
title: "HOmics preparing data"
author: "Lara Nonell"
date: "`r format(Sys.time(), '%d %b %Y')`"
output: 
  BiocStyle::html_document:
    number_sections: true
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Introduction

HOmics is a package to integrate omics data by capturing hierarchical omics relations in their association to phenotype. 
It contains two main functions, the generic HOmics function and HOmics.meth, to analyze specifically methylation of genes affecting to a phenotype. Details on how to precisely apply this latter function are available in the *HOmics gene methylation integration* vignette.

# Installation

HOmics needs the JAGS environment. JAGS (Just Another Gibbs Sampler), is a program for analysis of Bayesian hierarchical models using Markov Chain Monte Carlo (MCMC) simulation.
You install JAGS easily through this link: [http://mcmc-jags.sourceforge.net/](http://mcmc-jags.sourceforge.net/)

# Data

Input data for main function HOmics are the following:

- data.matrix: matrix with rownames the features and columns the samples.
- agg.matrix: matrix with colnames the features and columns the aggregation criteria, 0 for non pertenance.
- cond: response variable, usually a numerical factor with two levels representing the conditions to compare. If cond is a numerical vector (continuous response), a hiearchical linear regression model will be fit instead of the default hierarchical logistic regression model.
- z.matrix column of annotation that contains gene names (same annotations as gene.list). 
- cores: cores in case of parallelization. Default=1.

## Other

## Response variable, phenotype or condition

pheno.cond.col should be a variable  in the annotation data  associated to meth.data
usually a column in pData(meth.data)

Response variable can be continuous or categorical. 

### Categorigal responses

When response variable is categorical, a hierarchical logistic model will be performed.

In such cases, some considerations need to be taken.

JAGS needs a vector with 0 and 1, so it will be converted if it is not the case
The rules to convert the values are the following.

if it is already numerical, will be converted to a vector of 0's and 1's using the module 2 function denoted in R with %%

```{r cond1}
a <- c(2,1,1,1,2)
y <- a %% 2
y
```

if it is a character variable, it will be converted to a numerical factor, to be interpreted by JAGS

Remember when converting to factor in R, it always assigns the levels in alphabetical order.
So for instance vector a with 4 character categories, will finally converted to a numerical vector of 0's and 1's this way:

```{r cond2}
a <- c("z","t","b","z","t","b","z")
(a.f <- as.factor(a))
levels(a.f)
(a.f.n <- as.numeric(a.f))
(cond <- a.f.n %% 2)
```

### Continuous responses

When response variable is continuous, a hierarchical linear model will be performed.

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

# Annotations

GEO data from Illumina methylation array 27k (GPL...) and 450K (GPL...) contain two fields UCSC_RefGene_Name and UCSC_RefGene_Group but EPIC (GPL21145) does not contain such information. Annotation files
